Measuring precise radial velocities and cross-correlation function line-profile variations using a Skew Normal distribution

### 2018-05-24 manuscript.tex
Add files via upload

%\abstract
{\bf Abstract}\\
% Context, Aims, Methods, Results, Conclu (not mandatory)
Stellar activity is the main limitation to the detection of small-mass exoplanets using the radial-velocity (RV) technique. Stellar activity can be probed by measuring variations in the shape of the cross-correlation function (CCF) as a function of time. Those variations are measured using the different moments of the CCF, therefore measuring with the best precision those moments is crucial to de-correlate exoplanet signals from spurious RV signals originating from stellar activity.

We propose here to measure those moments using a Skew Normal (SN) distribution, which compared to the Normal distribution generally used, naturally provides an extra parameter to model the natural asymmetry of the CCF induced by convective blueshift.

We analyze 5 stars with different activity levels and different signal-to-noise ratio (SNR) levels. In each case, we compare the results obtained by fitting to the CCF respectively a Normal and a SN.  We also estimate rigorous errors for the different moments of the CCF using a bootstrap analysis.

The correlation between the RV's and the CCF asymmetry or the correlation between the RV's and the CCF width are always stronger when using the parameters derived from the SN in the case of real observations. Therefore the CCF asymmetry and the CCF width derived using a SN are more sensitive to stellar activity, which allows to probe with a better precision stellar rotational periods, in addition to characterize more precisely stellar activity signals. The precision on the estimated set of RV's, derived using the median of the fitted SN distributions are on average $10\%$ smaller than the uncertainties calculated on the mean of the Normal. In addition, the asymmetry parameter derived from the SN is on average $15\%$ better than the one calculated on the common Bisector Inverse Slope Span (BIS SPAN). We strongly encourage the use of the SN distribution rather than the Normal distribution, because this allows us to retrieve in one single fit the different moments of the CCF, because the derived moments better catch stellar activity signals and finally because the standard errors on the RV's and the the asymmetry parameter are smaller than the one estimated with the Normal fit.

@@

When working with radial velocities (RV's), the main limitation to the detection of small-mass exoplanets is not anymore the precision of the instruments used, but the different noises induced by the stars we are observing \citep[][]{Dumusque:2017aa}. Indeed, stellar oscillations, granulation phenomena and stellar activity all induce RV signals \citep[e.g.][]{Saar-1997b, Queloz-2001, Desort-2007, Dumusque-2011a, Dumusque-2016a} that are beyond the meter-per-second precision reached by the best high-resolution spectrographs. It is therefore mandatory to understand better stellar signals and to find ways to correct for them, if in the near future we want to detect or confirm an Earth-twin planet using the RV technique. This is even more true now that instrument like ESPRESSO \citep{Pepe-2014} and EXPRESS \citep{Fischer:2017aa} should have the stability to detect such signals. However, if solutions are not found to mitigate the impact of stellar activity, the detection or confirmation of potential Earth-twins will be extremely challenging.

@@

Among the different stellar signals we are aware of, the one that is the most difficult to characterize and to correct for is the signal induced by stellar activity. Stellar activity is responsible for creating magnetic regions on the surface of stars, and those regions change locally the temperature and the convection, inducing spurious RV's variations \citep[e.g.][]{Meunier-2010a, Dumusque-2014b}. In theory, it should be easy to differentiate between the pure Doppler-shift induced by a planet, that will shift the entire stellar spectrum, and stellar activity that modifies the shape of spectral lines and by doing so create a spurious shift of the stellar spectrum \citep{Saar-1997b,Hatzes-2002,Kurster2003,Lindegren-2003,Desort-2007,Lagrange-2010,Meunier-2010a,Dumusque-2014b}. However, on quiet GKM dwarfs, the main target for precise RV's measurements, stellar activity induce signals of a few \ms. This corresponds physically to variations smaller than 1/100th of a pixel on the detector.

To be able to measure such tiny variations, we average the information of all the line in the spectrum by cross correlating the stellar spectrum with a synthetic \citep[][]{Baranne-1996,Pepe-2002a} or an observed stellar template \citep[][]{Anglada-Escude-2012}. The result of this operation gives us the cross-correlation function (CCF). To measure the Doppler-shift between different spectra, and therefore get the RV's of a star as a function of time, we calculate the variations of the CCF barycenter. The barycenter is generally estimated by fitting a Normal distribution to the CCF, and taking its mean. Variations in line shape between different spectra, which indicate the presence of signals induced by stellar activity, is measured by analysing the different moments of the CCF. Generally, the width of the CCF is estimated using the FWHM of the fitted Normal distribution, and its asymmetry using the the bisector inverse slope span \citep[BIS SPAN,][]{Queloz-2001}.

@@

If a RV signal is induced by activity, generally a strong correlation will be observed between the RV and chromospheric activity indicators like \logrhk\,or H-$\alpha$ \citep[e.g.][]{Boisse-2009,Dumusque-2012,Robertson-2014}, but also between the RV and the FWHM of the CCF or its BIS SPAN \citep[][]{Queloz-2001,Boisse-2009,Queloz-2009,Dumusque-2016a}. It is therefore common now, that when fitting for a planetray signal, in addition to a Keplerian, the model includes in addition linear dependancies with the \logrhk, the FWHM and the BIS SPAN \citep[e.g.][]{Dumusque:2017aa,Feng:2017aa}. It is also common to add a Gaussian process to the model to account for the correlated noise induced by stellar activity. The hyperparameters of the Gaussian process are generally trained on the different activity indicators \citep[e.g.][]{Haywood-2014,Rajpaul-2015}. It is therefore essential for mitigating stellar activity to obtain activity indicators that are the most correlated to the RV's but also for which we can obtain the best precision.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%\subsection{Why using a SN distribution?}
In all the method described above, the RV and the FWHM are derived using a Normal distribution fitted to the CCF, and the asymmetry is estimated using another approach.

In this paper we propose to use a Skew Normal (SN) distribution to derive at the same time the RV, the FWHM and the asymmetry of the CCF, as this function includes a skewness parameter \citep[][]{Azzalini1985}. In addition, we know that for solar-type stars and cooler dwarfs, the bisector of the CCF as a "C"-shape due to convective blueshift \citep[e.g.][]{Dravins-1981, Gray-2009}. Therefore, fitting the CCF using a model that naturally includes an asymmetry, like the SN distribution, should give more precise results.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%\subsection{Outline of paper}
The paper is organized as follow. In Sec.~\ref{sec:2} we introduce the SN distribution and describe its applicability for modeling the CCF,  showing that the SN distribution is a better representation of observed CCF than a Normal distribution, and study how the SN parameters relate to the RV, FWHM and BIS SPAN of the CCF. In Sec.~\ref{sec:31} we present a simple model to correct from stellar activity which extends the ones proposed by \citet[e.g.][]{Dumusque:2017aa,Feng:2017aa}. In Sec.~\ref{sec:soap} we study the SN fit to the CCF using simulations coming from SOAP 2.0. In Sec.~\ref{sec:4}, we compare on real observations the sensitivity of the SN parameters to stellar activity with respect to other existing indicators. In Sec.~\ref{sec:5} we derive error bars for the different CCF moments and finally we discuss our results and conclude in Sec.~\ref{sec:discu} and Sec.~\ref{sec:conclu}.

@@

%The CCF represents the average shape of spectral lines and is expressed in flux as a function of radial-velocity.
The SN density shape is used to model the CCF.  In particular, using least-squares algorithm, we fit the following function:
%
\begin{eqnarray} \label{eq:3}
f_{CCF}(y_i) = \mathrm{C} - \mathrm{A} \times SN(y_i;\mu, \sigma^2, \gamma), \quad i = 1, \ldots, n
\end{eqnarray}
%
where C is an unknown offset fitting the continuum of the CCF, A is an unknown amplitude parameter and $y_1, \ldots, y_n$ are the set of RV's considered for the CCF. Note that the CCF is expressed in flux as a function of the lag of the cross-correlation template, expressed in RV.

@@

In the following of the paper, we define RV as the mean of the Normal distribution. Concerning the fit of the CCF using the SN, we present at first $2$ indicators that define the RV of the star: the mean of the SN, defined as SN mean RV and the median of the SN, defined as SN median RV (i.e. realling Eq.\eqref{eq:snmed}, $m$ = SN median RV). We will discuss advantages and limits for both these choices in Sec. \ref{sec:4} and Sec. \ref{sec:5}. For the width of the CCF, we use the FWHM of the Normal, which is $2\sqrt(2\ln2)\sigma$. The width of the SN, SN FWHM, is defined in the same way\footnote{Note that SN FWHM does not correspond to the width of the SN distribution at half maximum like in the Normal case.}. Being a Normal distribution symmetric, there is not such a parameter that evaluates the asymmetry of the distribution, so the BIS SPAN is used. The BIS SPAN will be compared to the asymmetric parameter $\gamma$ of the SN, named SN GAMMA. To test the strength of the correlation between the estimated RV's and the different indicators used to evaluate stellar activity, we calculated the Pearson correlation coefficient, which in its general form is defined as:
%
\begin{equation}
R (x,y)= \frac{cov(x,y)}{\sigma(x),\sigma(y)},
\label{eq:Pearson:corr}
\end{equation}
%
where $x$ and $y$ are two quantitative variables, $cov(x,y)$ indicates the covariance between $x$ and $y$, and $\sigma(x)$ and $\sigma(y)$ represent their standard deviations. 
A $p-$value for the statistical test having null hypothesis $H_{0}: R=0$ is provided, along with a $95\,\%$ confidence interval for $R$.

@@

Exoplanets will only produce a variation in RV's induced by a pure doppler shift of stellar spectra. Stellar activity and in particular the presence of active regions on the photosphere of the star, on the contrary, does not produce a blueshift or redshift of the spectra, but creates spurious RV's signal by modifying the shape of the spectral lines. To track these changes in the shape of the line profile, the general approach consists in using the FWHM, the BIS SPAN or the indicators introduced by \citep{Figueira-2013}, which provide an information on the average width and asymmetry of the CCF. A strong correlation between the estimated set of RV's and one or more of these parameters provides an indication that the RV's are affected by stellar activity signals.

Because of the spurious variations in RV's caused by active regions, a crucial step of the entire analysis is to correct the initially estimated RV's in order to retrieve a set of RV's containing ideally only the possibly pure doppler shift coming from the exoplanets. In order to reach this goal, we propose to consider a linear combination of the RV's having as covariates the amplitude parameter A (also known as contrast), the BIS SPAN (or $\gamma$), the FWHM (or SN FWHM) and the interaction between the BIS SPAN and the FWHM (or $\gamma$ and SN FWHM and the interaction between $\gamma$ and SN FWHM in the SN case). The parameters in parenthesis will be used when the SN fit is used. We propose the following function to correct the RV's from stellar activity:
%
\begin{equation}
RV_{\text{stellar activity}}= \beta_{0} + \beta_{1} A + \beta_{2} \gamma + \beta_{3} \text{SN FWHM} + \beta_{4} (\gamma  \text{SN FWHM})+\epsilon,
\label{eq:RV:correction}
\end{equation}
%
where $\beta_{0}$ is the intercept and $\epsilon$ is the vector of the errors with mean equal to $0$ and covariance matrix equal to $\sigma^{2}I$ ($I$ defined as the identity matrix).

While the parameters BIS SPAN (or $\gamma$) and FWHM (or SN FWHM) have already been introduced and used in \citep[e.g.][]{Dumusque:2017aa,Feng:2017aa}, by using Eq.~\eqref{eq:RV:correction} we propose also to include the contrast parameter $A$ and a variable that quantifies the interaction between BIS SPAN (or $\gamma$) and FWHM (or SN WHM). The inclusion of the contrast parameter $A$ is justified by the expected behavior of the CCF in presence of active regions. Fig.~\ref{fig:spotA} shows that the presence for instance of a spot produces also changes on the amplitude of the CCF and not only on its asymmetry or width.
%
\begin{figure*}[htbp]
   \centering
\includegraphics[height = 4in]{FTS_solar_CCFs_whyWeAddAlsoPARAM_A.pdf} 
   \caption{The presence of a spot on the photosphere of a star produces changes on the amplitude of the CCF (also known as contrast) and not only on its asymmetry or width. This very same plot has been originally presented by \citep{Dumusque-2014b} when introducing the Spot Oscillation And Planet (SOAP) 2.0 simulator.}
    \label{fig:spotA}
\end{figure*}
%
The reasons for including in Eq.~\eqref{eq:RV:correction} a variable that quantifies the interaction between BIS SPAN (or $\gamma$) and FWHM (or SN FWHM) will be better justified through the results of the examples presented in Sec.\ref{sec:soap}. We point out however that while the association between in this case BIS SPAN  (or $\gamma$) and FWHM (or SN FWHM) means that the values of one variable relate to the values of the other (since in this case we have two quantitative variables we talk about correlation), with the term interaction we mean that the effect that one variable has on the RV's is not constant. In particular the effect differs at different values of the other values. As a consequence of this, if two variables are interacting the may or may not be associated. In all the analyses presented in Sec. \ref{sec:soap} and Sec. \ref{sec:4} we found weak correlations between BIS SPAN (or $\gamma$) and FWHM (or SN FWHM) but the inclusion of the interaction term in Eq.~\eqref{eq:RV:correction} has shown nevertheless some utility in explaining spurious variations in RV's.

@@

As final remark we recall that, when working with a linear regression, there are several ways to select the variables to include in the model. While usually the stepwise technique takes place \citep{efroymson1960multiple,hocking1976biometrics}, the proposed function defined in Eq.~\eqref{eq:RV:correction} that corrects from stellar activity is the result of statistical and astronomical considerations. In particular we checked that the correlations between the proposed parameters were not approaching one: if it were the case, the matrix needed to calculate the estimates would have been singular, hence non invertible. This problem is known in statistic with the term multicollinearity and leads to unreasonable results and estimation of the standard errors associated to the parameters of interest. A detailed discussion of the topic can be found in the book by \citep{belsley1991conditioning}. In the literature there are examples of stars presenting a high correlation between the contrast parameter $A$ and the FWHM \citep[e.g.][]{Dumusque:2017aa,Feng:2017aa}. In the analyses of the presented work, we did not find any correlation between the contrast parameter $A$ and the FWHM (or between the contrast parameter $A$ and the SN FWHM) exceeding $0.66$ for all the analyzed stars and therefore the problem of multicollinearity is avoided. For the interpretation of the coefficients, we centered on the mean all the variables considered in the linear regression defined in Eq.~\eqref{eq:RV:correction}. Last, when proposing interactions between covariates, we found that only the one between the width and the asymmetry of the CCF is statistically helpful to explain spurious variations in RV's caused by stellar activity. Consequently, in the present work, the other interactions variables are not displayed.

@@

\begin{figure*}[htbp]
   \centering
\includegraphics[height = 4in]{SOAP_SPOT_Comparison_para_SN.pdf} 
   \caption{Evaluation of the correlation between the RVs and the asymmetry parameters when a spot is present on the photosphere of the star. In this case only the shape of the CCF changes as the spot moves, producing statistically significative correlations only between the RVs and the asymmetry parameter.}
    \label{fig:spot.corr}
\end{figure*}

@@

The last presented simulation example consists in having not only spurious variations in RV's caused by the presence of a spot on the photosphere of the star but also a pure doppler shift because of an artificially injected planet. The purpose of this example is to check if we are able to disentangle the two different sources of variations in the set of RV's.

Fig.~\ref{fig:spot.plus.planet} shows the results obtained when a spot is present on the photosphere of the star and a planet is injected. The planet, having an amplitude of 10 \ms, produces a pure doppler shift on the CCF, without further changing its shape.  In this case N mean RV seems to have the largest variations caused by the combined action of spot and planet. %The evaluation of the standard errors leads to the same conclusions reported for the two previous cases: SN mean RV has the largest related uncertainties and the standard errors associates with SN median RV are $10 \%$ smaller than the standard errors derived for RV.

@@

\begin{figure*}[htbp]
   \centering
\includegraphics[height = 4in]{RV_comparison_PLANET_SPOT.pdf} 
%\includegraphics[height = 2.5in]{RV_se_comparison_SPOT.pdf} 
 \caption{RV's changes as function of the orbital phase in the case in which a spot is present on the photosphere of the star and a planet is injected. N mean RV seems to have the largest variations caused by the combined action of spot and planet.}
   %\caption{(left) RV's changes as function of the orbital phase in the case in which a spot is present on the photosphere of the star and a planet is injected. N mean RV seems to have the largest variations caused by the combined action of spot and planet. (right) Evaluation of the standard errors corresponding to the defined RVs. The standard errors retrieved for SN median RV are $10 \%$ smaller than the standard errors derived for RV. SN mean RV has the largest related uncertainties.}
    \label{fig:spot.plus.planet}
\end{figure*}

\begin{figure*}[htbp]
   \centering
\includegraphics[height = 4in]{SOAP_SPOT_PLANET_Comparison_para_SN.pdf} 
   \caption{Evaluation of the correlation between the RVs and the asymmetry parameters when a spot is present on the photosphere of the star and a planet is injected.  In this case only the shape of the CCF changes as the spot moves, producing statistically significative correlations only between the RVs and the asymmetry parameter. The correlations between the RVs and the width parameter of the CCF is weaker than the previous case that considers only the presence of a spot on the photosphere of the star.}
    \label{fig:spot.plus.planet.corr}
\end{figure*}

In order to correct the estimated set of RV's from spurious variations caused by the spot, we need to generalize Eq.~\eqref{eq:RV:correction}, in order to consider also the possibly presence of an exoplanet. Hence, rather than just using Eq.~\eqref{eq:RV:correction} we add the following keplerian function, which represents the pure doppler shift of the exoplanet under the assumption of circular orbit:
%
\begin{equation}
RV_{\text{exoplanet}}= K \sin \left(\frac{2 \pi}{P} (t - t_{0})\right),
\label{eq:RV:correction.planet}
\end{equation}
%
where the amplitude $K$, the orbital period $P$ and the epoch at the periapsis $t_{0}$ are three unknown parameters that define the presence of the exoplanet. Overall, the variations in RV's of the star can be defined as
%
\begin{equation}
RV= RV_{\text{stellar activity}} + RV_{\text{exoplanet}}.
\label{eq:RV:correction.overall}
\end{equation}
%

>>>>

%\abstract
{\bf Abstract}\\
% Context, Aims, Methods, Results, Conclu (not mandatory)
Stellar activity is the main limitation to the detection of small-mass exoplanets using the radial-velocity (RV) technique. Stellar activity can be probed by measuring variations in the shape of the cross-correlation function (CCF) as a function of time. Those variations are measured using the different moments of the CCF. Therefore measuring with the best precision those moments is crucial to de-correlate exoplanet signals from spurious RV signals originating from stellar activity.

We propose here to measure those moments using a Skew Normal (SN) distribution, which compared to the Normal distribution generally used, automatically includes an extra parameter to model the natural asymmetry of the CCF induced by convective blueshift.

We analyze 5 stars with different activity levels and whose CCF's have different signal-to-noise ratio (SNR) levels. In each case, we compare the results obtained by fitting to the CCF respectively a Normal and a SN.  We also estimate rigorous errors for the different moments of the CCF using a bootstrap analysis.

The correlation between the RV's and the asymmetry of the CCF and the correlation between the RV's and the width of the CCF are always stronger when using the parameters derived from the SN in the case of real observations. Therefore the CCF asymmetry and the CCF width derived using a SN are more sensitive to stellar activity, which allows to probe with a better precision stellar rotational periods, in addition to characterize more precisely stellar activity signals. The precision on the estimated set of RV's, derived using the median of the fitted SN distributions are on average $10\%$ smaller than the uncertainties calculated on the mean of the Normal. In addition, the uncertainties related to the asymmetry parameter derived from the SN are on average $15\%$ smaller than the ones calculated on the common Bisector Inverse Slope Span (BIS SPAN). We strongly encourage the use of the SN distribution rather than the Normal distribution, because this allows us to retrieve in one single fit the different moments of the CCF, because the derived moments better catch stellar activity signals and finally because the standard errors on the RV's and the the asymmetry parameter are smaller than the one estimated with the Normal fit.

@@

When working with radial velocities (RV's), the main limitation to the detection of small-mass exoplanets is not anymore the precision of the instruments used, but the different noises induced by the stars we are observing \citep[][]{Dumusque:2017aa}. Indeed, stellar oscillations, granulation phenomena and stellar activity all induce RV signals \citep{Saar-1997b, Queloz-2001, Desort-2007, Dumusque-2011a, Dumusque-2016a} that are beyond the meter-per-second precision reached by the best high-resolution spectrographs. It is therefore mandatory to understand better stellar signals and to find ways to correct for them, if in the near future we want to detect or confirm an Earth-twin planet using the RV technique. This is even more true now that instrument like ESPRESSO \citep{Pepe-2014} and EXPRESS \citep{fischer2016state} should have the stability to detect such signals. However, if solutions are not found to mitigate the impact of stellar activity, the detection or confirmation of potential Earth-twins will be extremely challenging.

@@

Among the different stellar signals we are aware of, the one that is the most difficult to characterize and to correct for is the signal induced by stellar activity. Stellar activity is responsible for creating magnetic regions on the surface of stars, and those regions change locally the temperature and the convection, inducing spurious RV's variations \citep{Meunier-2010a, Dumusque-2014b}. In theory, it should be easy to differentiate between the pure Doppler-shift induced by a planet, that will shift the entire stellar spectrum and stellar activity, that modifies the shape of spectral lines and by doing so create a spurious shift of the stellar spectrum \citep{Saar-1997b,Hatzes-2002,Kurster2003,Lindegren-2003,Desort-2007,Lagrange-2010,Meunier-2010a,Dumusque-2014b}. However, on quiet GKM dwarfs, the main target for precise RV's measurements, stellar activity induce signals of a few \ms. This corresponds physically to variations smaller than 1/100th of a pixel on the detector.

To be able to measure such tiny variations, we average the information of all the lines in the spectrum by cross correlating the stellar spectrum with a synthetic \citep[][]{Baranne-1996,Pepe-2002a} or an observed stellar template \citep[][]{Anglada-Escude-2012}. The result of this operation gives us the cross-correlation function (CCF). To measure the Doppler-shift between different spectra and therefore to retrieve the RV's of a star as a function of time, we calculate the variations of the CCF barycenter. The barycenter is generally estimated by fitting a Normal distribution to the CCF and taking its mean. Variations in line shape between different spectra, which indicate the presence of signals induced by stellar activity, is measured by analysing the different moments of the CCF. Usually, the width of the CCF is estimated using the FWHM of the fitted Normal distribution, and its asymmetry using the the bisector inverse slope span \citep[BIS SPAN,][]{Queloz-2001}.

@@

If a RV signal is induced by activity, generally a strong correlation will be observed between the RV and chromospheric activity indicators like \logrhk\,or H-$\alpha$ \citep{Boisse-2009,Dumusque-2012,Robertson-2014}, but also between the RV and the FWHM of the CCF or its BIS SPAN \citep[][]{Queloz-2001,Boisse-2009,Queloz-2009,Dumusque-2016a}. It is therefore common now, that when fitting for a planetray signal, in addition to a Keplerian, the model includes in addition linear dependancies with the \logrhk, the FWHM and the BIS SPAN \citep{Dumusque:2017aa,Feng:2017aa}. It is also common to add a Gaussian process to the model to account for the correlated noise induced by stellar activity. The hyperparameters of the Gaussian process are generally trained on the different activity indicators \citep{Haywood-2014,Rajpaul-2015}. It is therefore essential for mitigating stellar activity to obtain activity indicators that are the most correlated to the RV's but also for which we can obtain the best precision.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%\subsection{Why using a SN distribution?}
In all the methods described above, the RV and the FWHM are derived using a Normal distribution fitted to the CCF, and the asymmetry is estimated using another approach.

In this paper we propose to use a Skew Normal (SN) distribution to derive at the same time the RV, the FWHM and the asymmetry of the CCF, as this function includes a skewness parameter \citep[][]{Azzalini1985}. In addition, we know that for solar-type stars and cooler dwarfs, the bisector of the CCF as a "C"-shape due to convective blueshift \citep{Dravins-1981, Gray-2009}. Therefore, fitting the CCF using a model that naturally includes an asymmetry, like the SN distribution, should give in principle more precise results.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%\subsection{Outline of paper}
The paper is organized as follow. In Sec.~\ref{sec:2} we introduce the SN distribution and describe its applicability for modeling the CCF,  showing that the SN distribution is a better representation of observed CCF than a Normal distribution, and study how the SN parameters relate to the RV, FWHM and BIS SPAN of the CCF. In Sec.~\ref{sec:31} we present a simple model to correct from stellar activity which extends the ones previously proposed \citep{Dumusque:2017aa,Feng:2017aa}. In Sec.~\ref{sec:soap} we study the SN fit to the CCF using simulations coming from SOAP 2.0. In Sec.~\ref{sec:4}, we compare on real observations the sensitivity of the SN parameters to stellar activity with respect to other existing indicators. In Sec.~\ref{sec:5} we derive error bars for the different CCF moments and finally we discuss our results and conclude in Sec.~\ref{sec:discu} and Sec.~\ref{sec:conclu}.

@@

%The CCF represents the average shape of spectral lines and is expressed in flux as a function of radial-velocity.
The SN density shape is used to model the CCF.  In particular, using least-squares algorithm, we fit the following function:
%
\begin{eqnarray} \label{eq:3}
f_{CCF}(y_i) = \mathrm{C} - \mathrm{A} \times SN(y_i;\mu, \sigma^2, \gamma), \quad i = 1, \ldots, n
\end{eqnarray}
%
where C is an unknown offset fitting the continuum of the CCF, A is an unknown amplitude parameter known with the term ``contrast'' and $y_1, \ldots, y_n$ are the set of RV's considered for the CCF. Note that the CCF is expressed in flux as a function of the lag of the cross-correlation template, expressed in RV.

@@

In the following of the paper, we define RV as the mean of the Normal distribution. Concerning the fit of the CCF using the SN, we present at first $2$ indicators that define the RV of the star: the mean of the SN, defined as SN mean RV and the median of the SN, defined as SN median RV (i.e. realling Eq.\eqref{eq:snmed}, $m$ = SN median RV). We will discuss advantages and limits for both these choices in Sec. \ref{sec:4} and Sec. \ref{sec:5}. For the width of the CCF, we use the FWHM of the Normal, which is $2\sqrt(2\ln2)\sigma$. The width of the SN, SN FWHM, is defined in the same way\footnote{Note that SN FWHM does not correspond to the width of the SN distribution at half maximum like in the Normal case.}. Being a Normal distribution symmetric, there is not such a parameter that evaluates the asymmetry of the distribution, so the BIS SPAN is used. The BIS SPAN will be compared to the asymmetric parameter $\gamma$ of the SN, known also as SN GAMMA. To test the strength of the correlation between the estimated RV's and the different indicators used to evaluate stellar activity, we calculated the Pearson correlation coefficient, which in its general form is defined as:
%
\begin{equation}
R (x,y)= \frac{cov(x,y)}{\sigma(x),\sigma(y)},
\label{eq:Pearson:corr}
\end{equation}
%
where $x$ and $y$ are two quantitative variables, $cov(x,y)$ indicates the covariance between $x$ and $y$, and $\sigma(x)$ and $\sigma(y)$ represent their standard deviations.  A $p-$value for the statistical test having null hypothesis $H_{0}: R=0$ is provided, along with a $95\%$ confidence interval for $R$ when needed.

@@

Exoplanets will only produce a variation in RV's induced by a pure Doppler-shift of stellar spectra. Stellar activity and in particular the presence of active regions on the photosphere of the star, on the contrary, does not produce a blueshift or redshift of the spectra, but creates spurious RV's signal by modifying the shape of the spectral lines. To track these changes in the shape of the line profile, the general approach consists in using the FWHM, the BIS SPAN or the indicators introduced by \citep{Figueira-2013}, which provide an information on the average width and asymmetry of the CCF. A strong correlation between the estimated set of RV's and one or more of these parameters provides an indication that the RV's are affected by stellar activity signals.

Because of the spurious variations in RV's caused by active regions, a crucial step of the entire analysis is to correct the initially estimated RV's in order to retrieve a set of RV's containing ideally only the possibly pure Doppler-shift coming from the exoplanets. In order to reach this goal, we propose to consider a linear combination of the RV's having as covariates the amplitude parameter A, the BIS SPAN (or $\gamma$), the FWHM (or SN FWHM) and the interaction between the BIS SPAN and the FWHM (or $\gamma$ and SN FWHM and the interaction between $\gamma$ and SN FWHM in the SN case). We propose the following function to correct the RV's from stellar activity:
%
\begin{equation}
RV_{\text{stellar activity}}= \beta_{0} + \beta_{1} A + \beta_{2} \gamma + \beta_{3} \text{SN FWHM} + \beta_{4} (\gamma  \text{SN FWHM})+\epsilon,
\label{eq:RV:correction}
\end{equation}
%
where $\beta_{0}$ is the intercept and $\epsilon$ is the vector of the errors with mean equal to $0$ and covariance matrix equal to $\sigma^{2}I$ ($I$ defined as the identity matrix).

While the parameters BIS SPAN and FWHM have already been introduced and used in \citep{Dumusque:2017aa,Feng:2017aa}, by using Eq.~\eqref{eq:RV:correction} we propose also to include the contrast parameter $A$ and a variable that quantifies the interaction between BIS SPAN (or $\gamma$) and FWHM (or SN WHM). The inclusion of the contrast parameter $A$ is justified by the expected behavior of the CCF in presence of active regions. Fig.~\ref{fig:spotA} shows that the presence for instance of a spot produces also changes in the amplitude of the CCF and not only on its asymmetry or width.
%
\begin{figure*}[htbp]
   \centering
\includegraphics[height = 4in]{FTS_solar_CCFs_whyWeAddAlsoPARAM_A.pdf} 
   \caption{The presence of a spot on the photosphere of a star produces changes on the amplitude of the CCF (also known as contrast) and not only on its asymmetry or width. This very same plot has been originally presented by \citep{Dumusque-2014b} when introducing the Spot Oscillation And Planet (SOAP) 2.0 simulator.}
    \label{fig:spotA}
\end{figure*}
%

The reasons for including in Eq.~\eqref{eq:RV:correction} a variable that quantifies the interaction between BIS SPAN (or $\gamma$) and FWHM (or SN FWHM) will be better justified through the results of the examples presented in Sec.\ref{sec:soap}. We point out however that while the association between in this case BIS SPAN  (or $\gamma$) and FWHM (or SN FWHM) means that the values of one variable relate to the values of the other (since in this case we have two quantitative variables we talk about correlation), with the term interaction we mean that the effect that one variable has on the RV's is not constant. In particular the effect differs at different values of the other values. %As a consequence of this, if two variables are interacting the may or may not be associated.

@@

As final remark we recall that, when working with a linear regression, there are several ways to select the variables to include in the model. While usually the stepwise technique takes place \citep{efroymson1960multiple,hocking1976biometrics}, the proposed function defined in Eq.~\eqref{eq:RV:correction} that corrects from stellar activity is the result of statistical and astronomical considerations. In particular we checked that the correlations between the proposed parameters were not approaching one: if it were the case, the matrix needed to calculate the estimates would have been singular, hence non invertible. This problem is known in statistic with the term multicollinearity. A detailed discussion of the topic can be found in the book by \citep{belsley1991}. In the literature there are examples of stars presenting a high correlation between the contrast parameter $A$ and the FWHM \citep{Dumusque:2017aa,Feng:2017aa}. In the analyses of the presented work, we did not find any correlation between the contrast parameter $A$ and the FWHM (or between the contrast parameter $A$ and the SN FWHM) exceeding $0.66$ for all the analyzed stars and therefore the problem of multicollinearity is avoided. For the interpretation of the coefficients, we centered on the mean all the variables considered in the linear regression defined in Eq.~\eqref{eq:RV:correction}. Last, when proposing interactions between covariates, we found that only the one between the width and the asymmetry of the CCF is statistically helpful to explain spurious variations in RV's caused by stellar activity. Consequently, in the present work, the interaction between other variables are not used as covariates.

@@

\begin{figure*}[htbp]
   \centering
\includegraphics[height = 4in]{SOAP_SPOT_Comparison_para_SN.pdf} 
   \caption{Evaluation of the correlation between the RV's and the asymmetry parameters when a spot is present on the photosphere of the star. In this case only the shape of the CCF changes as the spot moves, producing statistically significative correlations only between the RV's and the asymmetry parameter.}
    \label{fig:spot.corr}
\end{figure*}

@@

The last presented simulation example consists in having not only spurious variations in RV's caused by the presence of a spot on the photosphere of the star but also a pure Doppler-shift because of an artificially injected planet. The purpose of this example is to check if we are able to disentangle the two different sources of variations in the set of RV's.

Fig.~\ref{fig:spot.plus.planet} shows the results obtained when a spot is present on the photosphere of the star and a planet is injected. The planet, having an amplitude of 10 \ms, produces a pure Doppler-shift on the CCF, without further changing its shape.  In this case N mean RV seems to have the largest variations caused by the combined action of spot and planet. %The evaluation of the standard errors leads to the same conclusions reported for the two previous cases: SN mean RV has the largest related uncertainties and the standard errors associates with SN median RV are $10 \%$ smaller than the standard errors derived for RV.

@@

\begin{figure*}[htbp]
   \centering
\includegraphics[height = 4in]{RV_comparison_SPOT_PLANET.pdf} 
%\includegraphics[height = 2.5in]{RV_se_comparison_SPOT.pdf} 
 \caption{RV's changes as function of the orbital phase in the case in which a spot is present on the photosphere of the star and a planet is injected. N mean RV seems to have the largest variations caused by the combined action of spot and planet.}
   %\caption{(left) RV's changes as function of the orbital phase in the case in which a spot is present on the photosphere of the star and a planet is injected. N mean RV seems to have the largest variations caused by the combined action of spot and planet. (right) Evaluation of the standard errors corresponding to the defined RVs. The standard errors retrieved for SN median RV are $10 \%$ smaller than the standard errors derived for RV. SN mean RV has the largest related uncertainties.}
    \label{fig:spot.plus.planet}
\end{figure*}

\begin{figure*}[htbp]
   \centering
\includegraphics[height = 4in]{SOAP_SPOT_PLANET_Comparison_para_SN.pdf} 
   \caption{Evaluation of the correlation between the RV's and the asymmetry parameters when a spot is present on the photosphere of the star and a planet is injected.  In this case only the shape of the CCF changes as the spot moves, producing statistically significative correlations only between the RVs and the asymmetry parameter. The correlations between the RVs and the width parameter of the CCF is weaker than the previous case that considers only the presence of a spot on the photosphere of the star.}
    \label{fig:spot.plus.planet.corr}
\end{figure*}

In order to correct the estimated set of RV's from spurious variations caused by the spot, we need to generalize Eq.~\eqref{eq:RV:correction}, in order to consider also the possibly presence of an exoplanet. Hence, rather than just using Eq.~\eqref{eq:RV:correction} we add the following keplerian function, which represents the pure Doppler-shift of the exoplanet under the assumption of circular orbit:
%
\begin{equation}
RV_{\text{exoplanet}}= K \sin \left(\frac{2 \pi}{P} (t - t_{0})\right),
\label{eq:RV:correction.planet}
\end{equation}
%
where the amplitude $K$, the orbital period $P$ and the epoch at the periapsis $t_{0}$ are three unknown parameters that define the presence of the exoplanet. Overall, the variations in RV's of the star can be defined as
%
\begin{equation}
RV= RV_{\text{stellar activity}} + RV_{\text{exoplanet}}.
\label{eq:RV:correction.overall}
\end{equation}
%
